Bayesian Sparsification Methods for Deep Complex-valued Networks

### 2020-06-21 paper/paper.tex
replaced textit-s with emph

$, respectively, and $z$ is a \textit{proper} $\cplx$-Gaussian vector if $z$ and

@@

% setup and common experiment settings
For each dataset we conduct a set of experiments in which we try out all combinations of
model kinds ($\real$ or $\cplx$), sparsification methods (VD or ARD) and input features.
Where reasonable, we also study the effects of halving ($\tfrac12\cplx$) and doubling ($2\real$)
of the size of intermediate layers in each network \citep{monning_evaluation_2018}.
%
Each experiment is replicated \textit{five} times to take into account random initialization,
random order of the mini-batch during SGD, stochastic outputs of the intermediate layers,
and non-determinism of single-precision computations on GPUs.

Each network within every experiment passes through three successive stages of training:
pre-training (\textit{``dense''}), sparsification (\textit{``sparsify''}), and fine-tuning
(\textit{``fine-tune''}) (sec.~\ref{sub:staging}). The parameters and optimizer state are
passed between stages, except for last one, which inherits just the parameters.
%
Every experiment uses ADAM optimizer, resets the learning rate to $10^{-3}$ before each
stage, and applies global $\ell_2$-norm gradient clipping at $0.5$.

@@

At the \textit{``dense''} stage every network is fit ``as-is'', retaining its original
non-Bayesian architecture, and using only the likelihood term from \eqref{eq:sgvb_estimator} without
variational dropout.

During the \textit{``sparsify''} stage we make every layer Bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:vd_prior} (or sec.~\ref{ssub:ard_prior}) or its $\real$ counterpart.
We inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:sgvb_estimator} to study the attainable sparsity levels:
\begin{equation}  \label{eq:elbo_with_coef}
  - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
  + \frac1{M} \sum_{k=1}^M
      \log p_{\phi}(z_{i_k} \mid g(\varepsilon_k; \theta))
  \,.
  \tag{\ref{eq:sgvb_estimator}'}
\end{equation}
By varying $C$ we adjust the importance of the sparsifying prior in learning $q_\theta(W)$,
meaning that higher value yields more irrelevant parameters during inference and makes
layers sparser upon termination of this stage.
% RFC: citation needed fot this "trick". Cite $\beta$-VAE maybe?

% on compresssion
Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sec:variational_dropout}).
Since the $q_\theta$ factorizes into univariate distributions, a $\cplx$ or $\real$ parameter
is considered non-zero if $\log \alpha \leq \tau$ for $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$, or zero otherwise.

The \textit{``fine-tune''} stage reverts the network back to non-Bayesian layers, applies
sparsity masks to its parameters and proceeds with training the kept parameters as during
the ``dense'' stage. Unless specified otherwise the network is initialized to the mode
of the learnt $q_\theta(W)$.

@@

We conduct a moderately sized experiment on MNIST-like datasets of $28\times 28$ greyscale
images to study the performance-compression trade-off of the proposed $\cplx$-variational
dropout: MNIST \citep{lecun_gradient-based_1998}, KMNIST \citep{clanuwat_deep_2018},
EMNIST \citep{cohen_emnist_2017} and Fashion-MNIST \citep{xiao_fashion-mnist_2017}.
%
% dataset size
We deliberately use a {fixed} random subset of $10k$ images from the train split of each
dataset to fit the networks and measure the performance with classification \textit{accuracy}
score on the usual test split.
% Limiting the size was done primarily to test the regularization strength of compression
% and to facilitate quicker experiments.

% model architectures
% (fft) C - R - C / 2
% (raw) R - C - 2 * R
We consider two simple architectures for $\real$ and $\cplx$ networks in this experiment,
which have been chosen for the purpose of illustrating the compression and understanding
the effects of experiment parameters. \textit{TwoLayerDenseModel} is a wide dense ReLU
network $784 \to 4096 \to n_\mathtt{out}$, and \textit{SimpleConvModel} is a ReLU net with
two $2d$ $k_5 s_1$ convolutions with filters $20 \to 50$, two $k_2 s_2$ average pooling steps,
and a dense classifier network $800 \to 500 \to n_\mathtt{out}$. The number of intermediate
feature is as specified in $\real$ and $\cplx$ models, is halved in complex models tagged
$\tfrac12\cplx$, and doubled in real models labelled $2\real$.

@@

Each figure shows the compression-accuracy trade-off of a particular method and input
features for \textit{SimpleConvModel} and \textit{TwoLayerDenseModel} models for all four
of the studied datasets (described in the main text): EMNIST-Letters on the \textit{top-left},
KMNIST -- \textit{top-right}, Fashion MNIST -- \textit{bottom-left}, and MNIST on the
\textit{bottom-right}.
%
Figures \ref{fig:appendix__mnist-like__trade-off__ARD__fft}, \ref{fig:appendix__mnist-like__trade-off__VD__fft},
\ref{fig:appendix__mnist-like__trade-off__ARD__raw}, and \ref{fig:appendix__mnist-like__trade-off__VD__raw}
present $\real$ and $\cplx$ models with \textit{the same intermediate feature sizes}.

@@

%notation
By $e_i$ we denote the $i$-th real unit vector of dimensionality \textit{conforming} to the
matrix-vector expression it is used in. Let $\vec{M}$ denote \textit{row-major} flattening
of $M$ into a vector, i.e. in lexicographic order of its indices. $\diag{(\cdot)}$ embeds
$\cplx^n$ into $n\times n$ diagonal matrices, and $\otimes$ denotes the Kronecker product,
for which we note the following identities $
  \vec{P Q R} = (P \otimes R^\top) \vec{Q}
$, and $
  (P \otimes R) (C \otimes S) = P Q \otimes R S
$ \citep{petersen_matrix_2012}.
% It follows from the definition of the Kronecker product
% of $P$ and $R^\top$ and the row-major order vectorization.

If we assume a mean field $\cplx$-Gaussian approximate posterior for $
  W \in \cplx^{n\times m}
$, then
\begin{equation}  \label{eq:c-gauss-vi-general-vec}
  \vec{W}
    \sim \mathcal{N}^{\cplx}_{nm} \bigl(
      \vec{\mu}, \diag{\vec{\Sigma}}, \diag{\vec{C}}
    \bigr)
  \,,
\end{equation}
where mean, \textit{variance} and relation are given by matrices $
  \mu, \Sigma, C \in \cplx^{n\times m}
$ with $\Sigma_{ij} \geq 0$, and $
  \lvert C_{ij} \rvert^2 \leq \Sigma_{ij}
$. For any $x \in \cplx^m$ and $b\in \cplx^n$ $
  y = W x + b
    = (I_n \otimes x^\top) \vec{W} + b
$, whence the covariance and relation matrices of $y$ are
\begin{align}
  (I \otimes x^\top) \diag{\vec{\Sigma}} (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (I \otimes x^\top) e_i \otimes e_j
    %   \Sigma_{ij} (e_i \otimes e_j)^\top (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (e_i \otimes x^\top e_j)
    %   \Sigma_{ij} (e_i \otimes x^\top e_j)^{\hop}
    % = \sum_{ij} (e_i \otimes x_j) \Sigma_{ij} (e_i^\top \otimes \conj{x}_j)
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m \Sigma_{ij} x_j \conj{x}_j
    \,,  \label{eq:cn-gauss-lrt-cov} \\
  (I \otimes x^\top) \diag{\vec{C}} (I \otimes x^\top)^\top
    % = \sum_{ij} (e_i \otimes x_j) C_{ij} (e_i \otimes x_j)^\top
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m C_{ij} x_j^2
    \,.  \label{eq:cn-gauss-lrt-rel}
\end{align}
Since \eqref{eq:cn-gauss-lrt-cov} and \eqref{eq:cn-gauss-lrt-rel} are diagonal, $(y_i)_{i=1}^n$
are independent whence \eqref{eq:cn-affine} implies
\begin{equation}  \label{eq:cplx-gauss-trick-appendix}
  y_i
    \sim \mathcal{N}^{\cplx}
      \bigl(
        e_i^\top \mu x + b_i,
        \, \sum_{j=1}^m \Sigma_{ij} \lvert x_j \rvert^2,
        \, \sum_{j=1}^m C_{ij} x_j^2
      \bigr)
    \,.
\end{equation}
Hence \eqref{eq:cplx-gauss-trick} follows.

>>>>

$, respectively, and $z$ is a \emph{proper} $\cplx$-Gaussian vector if $z$ and

@@

% setup and common experiment settings
For each dataset we conduct a set of experiments in which we try out all combinations of
model kinds ($\real$ or $\cplx$), sparsification methods (VD or ARD) and input features.
Where reasonable, we also study the effects of halving ($\tfrac12\cplx$) and doubling ($2\real$)
of the size of intermediate layers in each network \citep{monning_evaluation_2018}.
%
Each experiment is replicated \emph{five} times to take into account random initialization,
random order of the mini-batch during SGD, stochastic outputs of the intermediate layers,
and non-determinism of single-precision computations on GPUs.

Each network within every experiment passes through three successive stages of training:
pre-training (\emph{``dense''}), sparsification (\emph{``sparsify''}), and fine-tuning
(\emph{``fine-tune''}) (sec.~\ref{sub:staging}). The parameters and optimizer state are
passed between stages, except for last one, which inherits just the parameters.
%
Every experiment uses ADAM optimizer, resets the learning rate to $10^{-3}$ before each
stage, and applies global $\ell_2$-norm gradient clipping at $0.5$.

@@

At the \emph{``dense''} stage every network is fit ``as-is'', retaining its original
non-Bayesian architecture, and using only the likelihood term from \eqref{eq:sgvb_estimator} without
variational dropout.

During the \emph{``sparsify''} stage we make every layer Bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:vd_prior} (or sec.~\ref{ssub:ard_prior}) or its $\real$ counterpart.
We inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:sgvb_estimator} to study the attainable sparsity levels:
\begin{equation}  \label{eq:elbo_with_coef}
  - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
  + \frac1{M} \sum_{k=1}^M
      \log p_{\phi}(z_{i_k} \mid g(\varepsilon_k; \theta))
  \,.
  \tag{\ref{eq:sgvb_estimator}'}
\end{equation}
By varying $C$ we adjust the importance of the sparsifying prior in learning $q_\theta(W)$,
meaning that higher value yields more irrelevant parameters during inference and makes
layers sparser upon termination of this stage.
% RFC: citation needed fot this "trick". Cite $\beta$-VAE maybe?

% on compresssion
Between \emph{``sparsify''} and \emph{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sec:variational_dropout}).
Since the $q_\theta$ factorizes into univariate distributions, a $\cplx$ or $\real$ parameter
is considered non-zero if $\log \alpha \leq \tau$ for $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$, or zero otherwise.

The \emph{``fine-tune''} stage reverts the network back to non-Bayesian layers, applies
sparsity masks to its parameters and proceeds with training the kept parameters as during
the ``dense'' stage. Unless specified otherwise the network is initialized to the mode
of the learnt $q_\theta(W)$.

@@

We conduct a moderately sized experiment on MNIST-like datasets of $28\times 28$ greyscale
images to study the performance-compression trade-off of the proposed $\cplx$-variational
dropout: MNIST \citep{lecun_gradient-based_1998}, KMNIST \citep{clanuwat_deep_2018},
EMNIST \citep{cohen_emnist_2017} and Fashion-MNIST \citep{xiao_fashion-mnist_2017}.
%
% dataset size
We deliberately use a {fixed} random subset of $10k$ images from the train split of each
dataset to fit the networks and measure the performance with classification \emph{accuracy}
score on the usual test split.
% Limiting the size was done primarily to test the regularization strength of compression
% and to facilitate quicker experiments.

% model architectures
% (fft) C - R - C / 2
% (raw) R - C - 2 * R
We consider two simple architectures for $\real$ and $\cplx$ networks in this experiment,
which have been chosen for the purpose of illustrating the compression and understanding
the effects of experiment parameters. \emph{TwoLayerDenseModel} is a wide dense ReLU
network $784 \to 4096 \to n_\mathtt{out}$, and \emph{SimpleConvModel} is a ReLU net with
two $2d$ $k_5 s_1$ convolutions with filters $20 \to 50$, two $k_2 s_2$ average pooling steps,
and a dense classifier network $800 \to 500 \to n_\mathtt{out}$. The number of intermediate
feature is as specified in $\real$ and $\cplx$ models, is halved in complex models tagged
$\tfrac12\cplx$, and doubled in real models labelled $2\real$.

@@

Each figure shows the compression-accuracy trade-off of a particular method and input
features for \emph{SimpleConvModel} and \emph{TwoLayerDenseModel} models for all four
of the studied datasets (described in the main text): EMNIST-Letters on the \emph{top-left},
KMNIST -- \emph{top-right}, Fashion MNIST -- \emph{bottom-left}, and MNIST on the
\emph{bottom-right}.
%
Figures \ref{fig:appendix__mnist-like__trade-off__ARD__fft}, \ref{fig:appendix__mnist-like__trade-off__VD__fft},
\ref{fig:appendix__mnist-like__trade-off__ARD__raw}, and \ref{fig:appendix__mnist-like__trade-off__VD__raw}
present $\real$ and $\cplx$ models with \emph{the same intermediate feature sizes}.

@@

%notation
By $e_i$ we denote the $i$-th real unit vector of dimensionality \emph{conforming} to the
matrix-vector expression it is used in. Let $\vec{M}$ denote \emph{row-major} flattening
of $M$ into a vector, i.e. in lexicographic order of its indices. $\diag{(\cdot)}$ embeds
$\cplx^n$ into $n\times n$ diagonal matrices, and $\otimes$ denotes the Kronecker product,
for which we note the following identities $
  \vec{P Q R} = (P \otimes R^\top) \vec{Q}
$, and $
  (P \otimes R) (C \otimes S) = P Q \otimes R S
$ \citep{petersen_matrix_2012}.
% It follows from the definition of the Kronecker product
% of $P$ and $R^\top$ and the row-major order vectorization.

If we assume a mean field $\cplx$-Gaussian approximate posterior for $
  W \in \cplx^{n\times m}
$, then
\begin{equation}  \label{eq:c-gauss-vi-general-vec}
  \vec{W}
    \sim \mathcal{N}^{\cplx}_{nm} \bigl(
      \vec{\mu}, \diag{\vec{\Sigma}}, \diag{\vec{C}}
    \bigr)
  \,,
\end{equation}
where mean, \emph{variance} and relation are given by matrices $
  \mu, \Sigma, C \in \cplx^{n\times m}
$ with $\Sigma_{ij} \geq 0$, and $
  \lvert C_{ij} \rvert^2 \leq \Sigma_{ij}
$. For any $x \in \cplx^m$ and $b\in \cplx^n$ $
  y = W x + b
    = (I_n \otimes x^\top) \vec{W} + b
$, whence the covariance and relation matrices of $y$ are
\begin{align}
  (I \otimes x^\top) \diag{\vec{\Sigma}} (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (I \otimes x^\top) e_i \otimes e_j
    %   \Sigma_{ij} (e_i \otimes e_j)^\top (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (e_i \otimes x^\top e_j)
    %   \Sigma_{ij} (e_i \otimes x^\top e_j)^{\hop}
    % = \sum_{ij} (e_i \otimes x_j) \Sigma_{ij} (e_i^\top \otimes \conj{x}_j)
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m \Sigma_{ij} x_j \conj{x}_j
    \,,  \label{eq:cn-gauss-lrt-cov} \\
  (I \otimes x^\top) \diag{\vec{C}} (I \otimes x^\top)^\top
    % = \sum_{ij} (e_i \otimes x_j) C_{ij} (e_i \otimes x_j)^\top
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m C_{ij} x_j^2
    \,.  \label{eq:cn-gauss-lrt-rel}
\end{align}
Since \eqref{eq:cn-gauss-lrt-cov} and \eqref{eq:cn-gauss-lrt-rel} are diagonal, $(y_i)_{i=1}^n$
are independent whence \eqref{eq:cn-affine} implies
\begin{equation}  \label{eq:cplx-gauss-trick-appendix}
  y_i
    \sim \mathcal{N}^{\cplx}
      \bigl(
        e_i^\top \mu x + b_i,
        \, \sum_{j=1}^m \Sigma_{ij} \lvert x_j \rvert^2,
        \, \sum_{j=1}^m C_{ij} x_j^2
      \bigr)
    \,.
\end{equation}
Hence \eqref{eq:cplx-gauss-trick} follows.
